A Decision Procedure for (Co)datatypes in SMT Solvers

### 2016-02-10 jour/jour.tex
more

Here are a few examples of legal specifications of (co)datatype families:
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{llist}_{\vthinspace\tau} & \,=\, \const{LNil} \,\mid\, \const{LCons}(%\const{lhead}{:}\;
      \tau,\: %\const{ltail}{:}\;
      \ty{llist}_{\vthinspace\tau}) \\%[-.5\jot]
      \keyw{datatype}\;\, \ty{tree}_{\vthinspace\tau} & \,=\, \const{Node}(\tau,\:\, \ty{forest}_{\vthinspace\tau}) \\[-\jot]
      \keyw{and}\;\,\, \ty{forest}_{\vthinspace\tau} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree}_{\vthinspace\tau},\:\, \ty{forest}_{\vthinspace\tau})
\end{aligned}
\]

@@

Datatypes and codatatypes share many basic properties. All properties
below are implicitly universally quantified and range over all $i$, $j$,
$\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & %\forall \bar x, \bar y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \quad\text{if $j \not= \jPrime$}}
  \\[-0.5\jot]
\text{Injectivity:}\quad
  & %\forall x_1,\ldots,x_{n_{i\negvthinspace j}} y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(y_1,\ldots,y_{n_{i\negvthinspace j}}) \medrightarrow x_k \teq y_k}
  \\[-0.5\jot]
\text{Exhaustiveness:}\quad
  & \smash{\const{d}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \const{d}_{im_i}(x)}
  \\[-0.5\jot]
\text{Selection:}\quad
  & \smash{\const{s}_{i\negvthinspace j}^{\,k}(\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}})) \teq x_k}
\end{aligned}
\]
%
Expressed in the algebraic jargon, exhaustiveness helps ensure that ``no
junk'' exists, whereas distinctness and injectivity guarantee that ``no
confusion'' can arise.
The result of selectors applied to the wrong
constructor is left completely unspecified.
%
%
%  * how to deal with "wrong" selectors, e.g.
%        hd(nil1) = hd(nil2)?
%      * leave them unspecified; hence hd(nil1) = hd(nil2) in some models,
%        and not in other models
%
Datatypes are additionally characterized by an induction principle for proving
a conjunction of properties $P_1,\ldots,P_{\ell}$ over arbitrary
values $v_1 : \delta_1, \ldots, v_\ell : \delta_\ell$:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{\strut$\bigwedge_{\vthinspace i=1}^\ell \bigwedge_{\vthinspace j=1}^{m_i}\, \forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\vthinspace k=1}^{ n_{i\negvthinspace j}}\, \mathcalx{IH}[x_k]\bigr) \medrightarrow P_i[\const C_{i\negvthinspace j}(x_1,\ldots,x_{n_{ij}})]$}
\UIC{\strut$\bigwedge_{\vthinspace i=1}^{\ell}\, P_i[v_i]$}
\DP
\end{aligned}
\]
The notation $\mathcalx{IH}[x]$
denotes either $P_{\iPrime}[x]$ if there exists some $\iPrime$ such that
the formula is type-correct or else $\top$ (truth).
The induction principle ensures that the interpretation of datatypes is standard.
For the natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction prohibits models that contain cyclic values---e.g.,
an $n$ such that $n \teq \const{S}(n)$---or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.

For codatatypes, the dual notion is called coinduction. It makes it possible to
derive the equality of pairs of codatatype values
$v_1, w_1 \mathbin: \delta_1,\,\ldots,\,v_\ell, w_\ell \mathbin: \delta_\ell$, based on
suitable coinduction witnesses $R_1,\ldots,R_\ell$:
%
\[
\begin{aligned}[t]
\text{Coinduction:}\quad
\AXC{\strut$\begin{gathered}\textstyle \smash{\bigwedge_{\vthinspace i=1}^\ell \, R_i[v_i, w_i]} \\[-\jot]\textstyle
\bigwedge_{\vthinspace i=1}^\ell\, \bigl(\forall v\; w.\;\, R_i[v, w] \medrightarrow
  \bigwedge_{\vthinspace j=1}^{m_i}\, \bigl(\const{d}_{\negvthinspace j}(v) \mathrel\land \const{d}_{\negvthinspace j}(w)
  \medrightarrow \bigwedge_{\vthinspace k=1}^{n_{i\negvthinspace j}}\, \const{s}_{i\negvthinspace j}^{\,k}(v) \sim \const{s}_{i\negvthinspace j}^{\,k}(w)\bigr)\bigr)
\end{gathered}$}
\UIC{\strut$\bigwedge_{\vthinspace i=1}^\ell \, v_i \teq w_i$}
\DP
\end{aligned}
\]
The notation $x \sim y$ stands for $R_{\iPrime}[x, y]$ if there exists some
$\iPrime$ such that the formula is type-correct or $x \teq y$ otherwise.
The second premise ensures that the coinduction witnesses are
bisimulations. The first premise and the conclusion capture the notion that
equality is the largest bisimulation on codatatypes. Thus, the coinduction
principle encodes a form of extensionality:\ Two values that yield the same
observations must be equal, where the observations are made through selectors
and discriminators.

In addition, codatatypes are guaranteed to contain all values corresponding to
infinite ground constructor terms.
In general, this cannot be captured by a
first-order axiomatization, since there may be uncountably many of them.
For example, $\ty{stream}_{\vthinspace\ty{int}}$ is isomorphic to the uncountable
function space $\ty{nat} \to \ty{int}$. Bj\o rner gives a rigorous treatment of
this aspect \cite{bjoerner-1998}.

@@

\begin{figure}[t!]
\normalsize
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t \in \tEc
}{
  \Ec := \Ec,\: t \teq t
}
\)
\rn{Refl}
\qquad
\(
\inferrule{
 t \teq u \in \Ec
}{
 \Ec := \Ec,\: u \teq t
}
\)
\rn{Sym}
\qquad
\(
\inferrule{
  s \teq t,\; t \teq u \in \Ec
}{
  \Ec := \Ec,\: s \teq u
}
\)
\rn{Trans}
\\[5\jot]
\(
\inferrule{
  \vec t \teq \vec u \in \Ec
  \betweenantes
  \const f( \vec t \,),\, \const f( \vec u ) \in \tEc
}{
  \Ec := \Ec,\: \const f( \vec t \,) \teq \const f( \vec u )
}
\)
\rn{Cong}
\qquad
\(
\inferrule{
  t \teq u,\; t \tneq u \in \Ec
}{
  \bot
}
\)
\rn{Conflict}
\\[5\jot]
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{C}( \vec u ) \in \Ec
}{
  \Ec := \Ec,\: \vec t \teq \vec u
}
\)
\rn{Inject}
\qquad
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{D}( \vec u ) \in \Ec
  \betweenantes
  \const{C} \not= \const{D}
}{
  \bot
}
\)
\rn{Clash}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for bidirectional closure%.
}
\label{fig:cc-rules}
\end{figure}

\begin{figure}[t!]
\vspace*{+6pt} %% TYPESETTING HACK
\normalsize
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
%  t\typ{\delta}
%  \betweenantes
  \delta \in \Data
  \betweenantes
  \Val \smash{\ec{t\typ{\delta}}} = \MU x.\; u
  \betweenantes
  x \in \FV( u )
}{
  \bot
}
\)
\rn{Acyclic}
\kern1.5em\( %%% TYPESETTING: should be \quad
\inferrule{
% t, u\typ{\delta}
% \betweenantes
 \delta \in \Codata
 \betweenantes
 \Val \smash{\ec{t\typ{\delta}}} \aequiv \Val \smash{\ec{u\typ{\delta}}}
}{
 \Ec := \Ec,\: t \teq u
}
\)
\rn{Unique}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for acyclicity and uniqueness%.
}
\label{fig:ab-rules}
\end{figure}

\begin{figure}[t!]
\vspace*{+6pt} %% TYPESETTING HACK
\normalsize
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t\typ{\delta} \in \tEc
  \betweenantes
  \Ctr^\delta = \{ \const{C}_1, \ldots, \const{C}_m \}
\\
  \bigl( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\delta \bigr)
  \text{ or }
  \bigl( \delta \in \Data \text{ and } %\left|
  \delta
  %\right|
  \text{ is finite} \bigr)
}{
  \Ec := \Ec,\: \DISC{1} %\ROR
  \ROR \cdots \ROR \Ec := \Ec,\: \DISC{m}
}
\)
\rn{Split}
\\[5\jot]
\(
\inferrule{
  t\typ{\delta}, u\typ{\delta} \in \tEc
  \betweenantes
  \delta \in \Codata
  \betweenantes
  \delta \text{ is a singleton}
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Single}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

>>>>

Here are a few examples of legal specifications of (co)datatype families:
\[\begin{aligned}[t]
      \keyw{datatype}\; \,\ty{list}_{\vthinspace\tau} & \,=\, \const{Nil} \,\mid\, \const{Cons}(\const{hd}{:}\;
      \tau,\: \const{tl}{:}\; \ty{list}_{\vthinspace\tau}) \\%[-.5\jot]
      \keyw{codatatype}\; \,\ty{llist}_{\vthinspace\tau} & \,=\, \const{LNil} \,\mid\, \const{LCons}(\const{lhd}{:}\;
      \tau,\: \const{ltl}{:}\;
      \ty{llist}_{\vthinspace\tau}) \\%[-.5\jot]
      \keyw{datatype}\;\, \ty{tree}_{\vthinspace\tau} & \,=\, \const{Node}(\tau,\:\, \ty{forest}_{\vthinspace\tau}) \\[-\jot]
      \keyw{and}\;\,\, \ty{forest}_{\vthinspace\tau} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree}_{\vthinspace\tau},\:\, \ty{forest}_{\vthinspace\tau})
\end{aligned}
\]

@@

Datatypes and codatatypes share many basic properties. All properties
below are implicitly universally quantified and range over all $i$, $j$,
$\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & %\forall \bar x, \bar y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \quad\text{if $j \not= \jPrime$}}
  \\[-0.5\jot]
\text{Injectivity:}\quad
  & %\forall x_1,\ldots,x_{n_{i\negvthinspace j}} y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(y_1,\ldots,y_{n_{i\negvthinspace j}}) \medrightarrow x_k \teq y_k}
  \\[-0.5\jot]
\text{Exhaustiveness:}\quad
  & \smash{\const{d}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \const{d}_{im_i}(x)}
  \\[-0.5\jot]
\text{Selection:}\quad
  & \smash{\const{s}_{i\negvthinspace j}^{\,k}(\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}})) \teq x_k}
\end{aligned}
\]
%
Expressed in the algebraic jargon, exhaustiveness helps ensure that ``no
junk'' exists, whereas distinctness and injectivity guarantee that ``no
confusion'' can arise.
The result of selectors applied to the wrong
constructor is left completely unspecified.
%
%
%  * how to deal with "wrong" selectors, e.g.
%        hd(nil1) = hd(nil2)?
%      * leave them unspecified; hence hd(nil1) = hd(nil2) in some models,
%        and not in other models
%
Datatypes are additionally characterized by an induction principle for proving
a conjunction of properties $P_1,\ldots,P_{\ell}$ over arbitrary
values $v_1 : \delta_1, \ldots, v_\ell : \delta_\ell$:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{\strut$\bigwedge_{\vthinspace i=1}^\ell \bigwedge_{\vthinspace j=1}^{m_i}\, \forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\vthinspace k=1}^{ n_{i\negvthinspace j}}\, \mathcalx{IH}[x_k]\bigr) \medrightarrow P_i[\const C_{i\negvthinspace j}(x_1,\ldots,x_{n_{ij}})]$}
\UIC{\strut$\bigwedge_{\vthinspace i=1}^{\ell}\, P_i[v_i]$}
\DP
\end{aligned}
\]
The notation $\mathcalx{IH}[x]$
denotes either $P_{\iPrime}[x]$ if there exists some $\iPrime$ such that
the formula is type-correct or else $\top$ (truth).
The induction principle ensures that the interpretation of datatypes is standard.
For the natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction prohibits models that contain infinite values
$\const{S}(\const{S}(\ldots))$.

For codatatypes, the dual notion is called coinduction. It makes it possible to
derive the equality of pairs of codatatype values
$v_1, w_1 \mathbin: \delta_1,\,\ldots,\,v_\ell, w_\ell \mathbin: \delta_\ell$, based on
suitable coinduction witnesses $R_1,\ldots,R_\ell$:
%
\[
\begin{aligned}[t]
\text{Coinduction:}\quad
\AXC{\strut$\begin{gathered}\textstyle \smash{\bigwedge_{\vthinspace i=1}^\ell \, R_i[v_i, w_i]} \\[-\jot]\textstyle
\bigwedge_{\vthinspace i=1}^\ell\, \bigl(\forall x\; y.\;\, R_i[x, y] \medrightarrow
  \bigwedge_{\vthinspace j=1}^{m_i}\, \bigl(\const{d}_{\negvthinspace j}(x) \medrightarrow \const{d}_{\negvthinspace j}(y)
  \mathrel\land \bigwedge_{\vthinspace k=1}^{n_{i\negvthinspace j}}\, \const{s}_{i\negvthinspace j}^{\,k}(x) \sim \const{s}_{i\negvthinspace j}^{\,k}(y)\bigr)\bigr)
\end{gathered}$}
\UIC{\strut$\bigwedge_{\vthinspace i=1}^\ell \, v_i \teq w_i$}
\DP
\end{aligned}
\]
The notation $x \sim y$ stands for $R_{\iPrime}[x, y]$ if there exists some
$\iPrime$ such that the formula is type-correct or $x \teq y$ otherwise.
The second premise ensures that the coinduction witnesses are
bisimulations. The first premise and the conclusion capture the notion that
equality is the largest bisimulation on codatatypes. Thus, the coinduction
principle encodes a form of extensionality:\ Two values that yield the same
observations must be equal, where the observations are made through selectors
and discriminators.

\begin{examplex}
The induction principle for $\ty{list}_{\vthinspace\tau}$ is given below:
\[
\AXC{\strut$P[\const{Nil}]$}
  \AXC{\strut$\forall x,\, \mathit{xs}.\;\vthinspace P[\mathit{xs}] \medrightarrow P[\const{Cons}(x, \mathit{xs})]$}
\BIC{\strut$P[vs]$}
\DP
\]
Assuming that $\const{lnull}$ and its negation are the discriminators
associated with \const{LNil} and \const{LCons},
the coinduction principle for $\ty{llist}_{\vthinspace\tau}$ is
\[
\AXC{\strut$R[\mathit{vs},\, \mathit{ws}]$}
  \AXC{$\phantom{\forall \mathit{xs}\; \mathit{ys}.\;\,} \!\begin{aligned}[b]
      \llap{\ensuremath{\forall \mathit{xs}\; \mathit{ys}.\;\,}}
    & R[\mathit{xs}, \mathit{ys}]
      \medrightarrow
      \bigl(\const{lnull}(\mathit{xs}) \medrightarrow \const{lnull}(\mathit{ys})\bigr)
\\[-\jot]      
    & \kern2em{\mathrel\land}\;
      \bigl(\lnot\;\const{lnull}(\mathit{xs}) \medrightarrow \lnot\;\const{lnull}(\mathit{ys})
        \mathrel\land \const{hd}(\mathit{xs}) \teq \const{hd}(\mathit{ys})
        \mathrel\land R[\const{tl}(\mathit{xs}), \const{tl}(\mathit{ys})]\bigr)
\end{aligned}$}
\BIC{\strut$\mathit{vs} \teq \mathit{ws}$}
\DP
\]
\xend
\end{examplex}

Codatatypes are guaranteed to contain all values corresponding to
infinite ground constructor terms.
In general, this cannot be captured by a
first-order axiomatization, since there may be uncountably many of them.
For example, $\ty{stream}_{\vthinspace\ty{int}}$ is isomorphic to the uncountable
function space $\ty{nat} \to \ty{int}$. Bj\o rner gives a rigorous treatment of
this aspect \cite{bjoerner-1998}.

@@

\begin{figure}[t!]
\normalsize
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t \in \tEc
}{
  t \teq t \in \Ec
}
\)
\rn{Refl}
\qquad
\(
\inferrule{
 t \teq u \in \Ec
}{
 u \teq t \in \Ec
}
\)
\rn{Sym}
\qquad
\(
\inferrule{
  s \teq t,\; t \teq u \in \Ec
}{
  s \teq u \in \Ec
}
\)
\rn{Trans}
\\[5\jot]
\(
\inferrule{
  \vec t \teq \vec u \in \Ec
  \betweenantes
  \const f( \vec t \,),\, \const f( \vec u ) \in \tEc
}{
  \const f( \vec t \,) \teq \const f( \vec u ) \in \Ec
}
\)
\rn{Cong}
\qquad
\(
\inferrule{
  t \teq u,\; t \tneq u \in \Ec
}{
  \bot
}
\)
\rn{Conflict}
\\[5\jot]
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{C}( \vec u ) \in \Ec
}{
  \vec t \teq \vec u \in \Ec
}
\)
\rn{Inject}
\qquad
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{D}( \vec u ) \in \Ec
  \betweenantes
  \const{C} \not= \const{D}
}{
  \bot
}
\)
\rn{Clash}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for bidirectional closure%.
}
\label{fig:cc-rules}
\end{figure}

\begin{figure}[t!]
\vspace*{+6pt} %% TYPESETTING HACK
\normalsize
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
%  t\typ{\delta}
%  \betweenantes
  \delta \in \Data
  \betweenantes
  \Val \smash{\ec{t\typ{\delta}}} = \MU x.\; u
  \betweenantes
  x \in \FV( u )
}{
  \bot
}
\)
\rn{Acyclic}
\kern1.5em\( %%% TYPESETTING: should be \quad
\inferrule{
% t, u\typ{\delta}
% \betweenantes
 \delta \in \Codata
 \betweenantes
 \Val \smash{\ec{t\typ{\delta}}} \aequiv \Val \smash{\ec{u\typ{\delta}}}
}{
 t \teq u \in \Ec
}
\)
\rn{Unique}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for acyclicity and uniqueness%.
}
\label{fig:ab-rules}
\end{figure}

\begin{figure}[t!]
\vspace*{+6pt} %% TYPESETTING HACK
\normalsize
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t\typ{\delta} \in \tEc
  \betweenantes
  \Ctr^\delta = \{ \const{C}_1, \ldots, \const{C}_m \}
\\
  \bigl( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\delta \bigr)
  \text{ or }
  \bigl( \delta \in \Data \text{ and } %\left|
  \delta
  %\right|
  \text{ is finite} \bigr)
}{
  \DISC{1} \in \Ec \ROR \cdots \ROR \DISC{m} \in \Ec
}
\)
\rn{Split}
\\[5\jot]
\(
\inferrule{
  t\typ{\delta}, u\typ{\delta} \in \tEc
  \betweenantes
  \delta \in \Codata
  \betweenantes
  \delta \text{ is a singleton}
}{
  t \teq u \in \Ec
}
\)
\rn{Single}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}
